FoRTE baseline

### 2019-01-31 analysis/paper/ed-notes.Rmd
PAPER: Add notes on finite canopy radius in ED

Crown area used in some of the following places:

- Canopy structural dynamics (`dynamics/canopy_struct_dynamics.f90:L470`)
- Plant hydrology (`dynamics/plant_hydro.f90:L145,204,240,447,626`)

@@

```
ca = biCa[pft] * dbh ^ b2Ca[pft]
```

But, note that crown area is restricted such that "Local LAI / Crown area" is never less than one (i.e you cannot have more crown area than leaf area index -- your maximum crown area is the leaf area index):

@@

Crown area set or calculated in:

- `dynamics/strcutral_growth.f90:L669,929`

>>>>

### Calculating crown area

@@

```
ca = b1Ca[pft] * dbh ^ b2Ca[pft]
```

Note that crown area is restricted such that "Local LAI / Crown area" is never less than one (i.e you cannot have more crown area than leaf area index -- your maximum crown area is the leaf area index):

@@

The allometry coefficients `b1Ca` and `b2Ca` have the following defaults (from `init/ed_params.f90:L3055`):

- `b1Ca`
    - 5:13 (C3 grass and temperate trees) -- `2.490154` (constant)
    - All others -- `exp(-1.853) * exp(b1Ht(pft)) ** 1.888`; log-linear relationship with height allometry
- `b2Ca`
    - 5:13 -- `0.8068806` (constant)
    - All others -- `b2Ht(pft) * 1.888`; linear relationship with height allometry

Note that only tropical PFTs are affected by the allometry model (`IALLOM`); coefficient adjustments are wrapped in `if (is_tropical(pft))`.

If two cohorts are fused, their crown areas are summed, up to a maximum of 1 (in `utils/fuse_fiss_utils.f90:L2932`).

### Radiative transfer

Crown area array (for each cohort) is passed to the various radiative transfer models (see `ICANRAD` parameter).
Inside the RTMs, it is the `cai` variable.

The following sections focus on the contributions of finite canopy radius to radiative transfer calculations.
For the full descriptions, consult the "Canopy radiative transfer" module section.

#### Two-stream

In the two-stream model, crown area is used to correct `mu_bar` (average inverse optical depth per unit leaf and stem area) for crown area (resulting variable `mu`):

```
mu = -etai / log(1 - cai + cai * exp(-etai / (cai * mu_bar)))
```

where `etai` is the total area index (TAI = LAI * clumping factor + WAI) of the layer.

(If `cai = 1`, this simplifies to just `mu_bar`).

In the longwave (thermal), crown area is also used as a linear scaling of the blackbody emissivity.

In the shortwave two-stream model, the crown area correction also happens for direct radiation (`mu0`):

```
mu0 = -etai / log(1 - cai + cai * exp((-proj_area * etai) / (cai * czen)))
```

where `czen` is the cosine of the solar zenith angle (radians) and `proj_area` is the projected area, calculated as a function of leaf orientation angle (`orient`):

```
phi1 = 0.5 - orient * (0.633 + 0.33 * orient)
phi2 = 0.877 * (1 - 2 * phi1)
proj_area = phi1 + phi2 * czen
```

These calculations follow CLM 4.5 [@clm45_note] (section 3.1, eq. 3.3), which in turn is based on @sellers_1985_canopy.
This is described in more detail in the canopy radiative transfer section.

#### Multiple-scatter

In the multiple-scatter model [@zhao_2005_multiple; @zhao_2006_modeling], crown area is used calculate the exposed total area index (TAI) of each layer (`locetai`):

```
locetai = etai / cai
```

Crown area is also used to adjust the layer transmittance coefficient for beam radiation (`tau_beam`):

```
lambda = proj_area / mu
tau_beam = (1 - cai) + cai * exp(-lambda * locetai)
```

A similar correction is applied to diffuse radiation, but the contribution from each hemisphere has to be integrated.
This uses the analytical solution for that integral.

```
d1 = phi1 * locetai
d2 = phi2 * locetai
tau_diff = (1 - cai) - cai * exp(-d1 * d2) * (d1^2 * exp(d1) * eifun(-d1) + (d1 - 1))
```

where `eifun` is the exponential integral function.

### Plant hydraulics

Source: `dynamcs/plant_hydro.f90`.

Subroutine `plant_hydro_driver`:
Crown area for each cohort calculated as `crown_area / nplant`, then passed to `calc_plant_water_flux` (as variable `crown_area`).

Calculate the root area index (RAI) at each depth:

```
RAI = broot_d * SRA * root_frac / (4 * crown_area)
```

where SRA is the PFT specific root area parameter, and `root_frac` is the fraction of roots in that layer (function of PFT parameter `root_beta` and depth).

Calculate the soil-root water conductance at each depth:

```
gw_cond = (4 * crown_area) * soil_cond * sqrt(RAI) / (pi * dslz)
```

where `soil_cond` is soil conductivity at that depth and `dslz` is the soil layer thickness.
This is based on @katul_2003_relationship.

### Canopy turbulence

In file `dynamics/canopy_struct_dynamics.f90`, subroutine `canopy_turbulence`.

Crown area used in calculation of the wind profile extinction coefficients.

```
#-- dynamics/canopy_struct_dynamics.f90:L470 --
extinct_full = crown_area * exp(-0.25 * lai / crown_area) + (1 - crown_area)
extinct_full = crown_area * exp(-0.5 * lai / crown_area) + (1 - crown_area)
veg_wind = max(ugbmin, uh[i - 1] * extinct_half)
uh[i] = uh[i - 1] * extinct_full
```

Note that there is a copy of the same calculation performed in double precision (subroutine `canopy_turbulence8`) later in the file.

Crown area is also used to calculate the open canopy fraction of each cohort by subroutine `update_patch_derived_props` (file `utils/update_derived_props.f90:L43`).
That subroutine is in turn called by subroutine `structural_growth_eq_0` (file `dynamics/structural_growth.f90`).
This open canopy fraction is used to calculate the canopy roughness (`rough`).

```
rough = snow_rough * snowfac_can +
    (soil_rough * opencan_frac + veg_rough * (1 - opencan_frac)) *
    (1 - snowfac_can)
```

Open canopy fraction is also used to calculate the net ground conductance (`ggnet`):

```
ggnet = ggbare * ggveg / (ggveg + ggbare * (1 - opencan_frac))
```

### 2019-01-31 analysis/paper/paper.Rmd
PAPER: Add notes on finite canopy radius in ED

- Finite canopy radius (`CROWN_MOD`)
    - Options:
        - (0; default): Complete shading
        - (1): Finite canopy radius (Dietze 2008)
    - See current hypotheses
    - This model is based on @dietze_2008_capturing
- Nitrogen limitation of photosynthesis (`N_PLANT_LIM`)
    - Options:
        - (0, default): No limitation
        - (1) Limitation based on internal model (TODO: what is it?)
    - Turning this on _should_ favor mid-successional species, and substantially increase uncertainty.
- Nitrogen limitation of decomposition (`N_DECOMP_LIM`)
    - Options:
        - (0, default): No limitation
        - (1) Limitation based on internal model (TODO: what is it?)
    - Same as above

@@

@@

        - (1) Arrhenius, `Gamma*` based on Michaelis-Mentel coefficients for CO2 and O2, as in Farquhar et al. (1980) (TODO: REF)
        - (2; default) Q10 equations and parameters of Collatz et al. (1991), with MM coefficients. High/low temperature correction in Moorcroft et al. (2001)
        - (3) As (2), but `Gamma*` as in Farquhar et al. (1980) and CLM.

@@

- Allometry (`IALLOM`) -- Seems to mostly affect tropical PFTs (TODO: Check that this is the case).
- Grass (`IGRASS`) -- I'm not running with any grass PFTs.
- Phenology (`IPHEN_SCHEME`) -- Differences affect grasses and tropical species. Conifers are always evergreen, and hardwoods are always cold-deciduous (need to check that schemes are the same).
- Seed dispersal (`REPRO_SCHEME`) -- Only seems to matter for multi-site runs, with some toggles for tropical (drought-deciduous) species.
- Hydrodynamics (`PLANT_HYDRO_SCHEME`) -- Seems mostly tropics-focused? Look into the Xu et al. (2016; New Phyt.) paper.
- Fire disturbance (`INCLUDE_FIRE`) -- We know there weren't any fires at UMBS during this time, and more generally are interested in stability.
    - This could potentially be fun to play with in a later paper about disturbance regimes. Potentially, with Jackie Matthes?
- Anthropogenic disturbances (`IANTH_DISTURB`) -- We are currently looking at stable state, without disturbances. But, since this pulls values from a dataset, this could be one way to apply the FoRTE treatments.

>>>>

- Finite canopy radius (`CROWN_MOD`)
    - Options:
        - (0; default): Complete shading
        - (1): Finite canopy radius (Dietze 2008)
    - See current hypotheses
    - This model is based on crown area allometries from @dietze_2008_capturing
- Nitrogen limitation of photosynthesis (`N_PLANT_LIM`)
    - Options:
        - (0, default): No limitation
        - (1) Limitation based on internal model (TODO: what is it?)
    - Turning this on _should_ favor mid-successional species, and substantially increase uncertainty.
- Nitrogen limitation of decomposition (`N_DECOMP_LIM`)
    - Options:
        - (0, default): No limitation
        - (1) Limitation based on internal model (TODO: what is it?)
    - Same as above

@@

- Canopy radiative transfer model (`ICANRAD`)
        - (0, deprecated?) Original two-stream model (Medvigy 2006)
        - (1) Multiple-scattering model [@zhao_2006_modeling; @zhao_2005_multiple]
        - (2, default) Updated two-stream model [@liou_2002]
    - In theory, multiple-scattering model is more detailed and should give "better" results. Open question as to how that affects predictions.
    - Good candidate for exploratory analysis. I already implemented the two-stream model in R -- could be a good idea to implement the multi-scatter model as well and compare their radiation profiles.
- Allometry (`IALLOM`)
        - (0, default): Original ED2
        - (1)
            - Structural biomass parameterized so total AGB similar to @baker_2004_variation, eq. 2. `Balive` similar to ED2.
            - Root depth configured so root depth is 5m for canopy trees and 0.5 m for grasses/seedlings
            - Crown area as in @poorter_2006_architecture, with maximum crown area
        - (2) Like (1), but...
            - Height to DBH allometry from @poorter_2006_architecture
            - `Balive` has extra allometric equations (see @cole_2006_allometric and @calvo-alvarado_2008_allometric)
        - (3) Like (2), but tropical leaf biomass estimated from @lescure_1983_la, and PFT differences in leaf biomass based on SLA rather than wood density. Also, split leaf allometry into small (sapling) and large (tree) classes, as in @cole_2006_allometric

@@

        - (1) Arrhenius, `Gamma*` based on Michaelis-Mentel coefficients for CO2 and O2, as in @farquhar_1980_biochemical
        - (2; default) Q10 equations and parameters of @collatz_1991_physiological, with MM coefficients. High/low temperature correction in @moorcroft_2001_method
        - (3) As (2), but `Gamma*` as in @farquhar_1980_biochemical and CLM.

@@

- Grass (`IGRASS`) -- I'm not running with any grass PFTs.
- Phenology (`IPHEN_SCHEME`) -- Differences affect grasses and tropical species. Conifers are always evergreen, and hardwoods are always cold-deciduous (need to check that schemes are the same).
- Seed dispersal (`REPRO_SCHEME`) -- Only seems to matter for multi-site runs, with some toggles for tropical (drought-deciduous) species.
- Hydrodynamics (`PLANT_HYDRO_SCHEME`) -- Seems mostly tropics-focused? Look into the Xu et al. (2016; New Phyt.) paper.
- Fire disturbance (`INCLUDE_FIRE`) -- We know there weren't any fires at UMBS during this time, and more generally are interested in stability.
    - This could potentially be fun to play with in a later paper about disturbance regimes. Potentially, with Jackie Matthes?
- Anthropogenic disturbances (`IANTH_DISTURB`) -- We are currently looking at stable state, without disturbances. But, since this pulls values from a dataset, this could be one way to apply the FoRTE treatments.

### Additional references

@katul_2003_relationship
